%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  Bachelor's/Master's Thesis       First Name      Last Name       Date
%  "Title"
%  Machine Learning and Data Analytics Lab, FAU Erlangen-Nuernberg
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass{maddoc}
% ++ es werden keine underfull hboxes als Fehler ausgegeben,
%    da das ja nur heißt, dass die Seite noch nicht ganz voll ist
\hbadness=10000

% load the bibliography file
\addbibresource{bibliography.bib}

% Nomenclature
%\usepackage[english]{nomencl}
%\makenomenclature

\usepackage{graphicx}
\usepackage{eurosym}
\usepackage{hhline} % for more flexible horizontal lines in tables
\usepackage{longtable,ltcaption}
\usepackage{minted}
\usepackage{caption}
\usepackage[htt]{hyphenat}
\definecolor{LightGray}{gray}{0.9}

% Ensure that lualatex looks very similar to pdflatex
\setmainfont{texgyretermes}[
  UprightFont = *-regular ,
  BoldFont = *-bold ,
  ItalicFont = *-italic ,
  BoldItalicFont = *-bolditalic ,
  Extension = .otf ,
  Scale = 1.0 ]
\defaultfontfeatures{Ligatures=TeX}

\input{config}
\input{commands}

\pagenumbering{roman}

\begin{document}
\begin{center}
\bfseries
Übersicht
\normalfont
\end{center}

Datenräume sind ein Datenintegrationskonzept, bei dem die Daten nicht in zentralen Datensilos gespeichert werden, sondern an der Quelle gespeichert bleiben.
Einzelne Föderationen werden eingerichtet, um die gemeinsame Nutzung von Daten zu ermöglichen, wobei die Datenkonsumenten über interoperable Standards mit den Datenanbietern verbunden sind.

Gaia-X zielt darauf ab, Datenräume in einer \textit{interoperablen}, \textit{transparenten}, \textit{offenen} und \textit{sicheren} Umgebung zu schaffen, die die \textit{Datenrechte} und die \textit{EU-Vorschriften} respektiert.
Gaia-X Credentials sind ein zentrales Konzept des Ökosystems.
Sie speichern Beschreibungen aller Teilnehmer und logischen Objekte in einer wohldefinierten und überprüfbaren Weise.
Die EU vertraut auf das Projekt, um die Abhängigkeit von ausländischen Tech-Giganten zu verringern und die Innovation in der digitalen Wirtschaft zu fördern.

In der Arbeit wird die Machbarkeit des Austauschs medizinischer Studiendaten in Gaia-X-Datenräumen untersucht.
Die Arbeit umfasst eine Literaturübersicht über Datenräume und verwandte Themen.

Die Hauptmethode der Evaluierung ist die Entwicklung eines Datenaustauschmoduls für die bestehende Plattform zur Speicherung medizinischer Studiendaten---Carecentive---unter Verwendung von Gaia-X-Komponenten und -Spezifikationen.
Das Austauschmodul dient als Proof of Concept und wird später zur Durchführung von Experimenten verwendet, um den Arbeitszustand der Gaia-X-Dienste und -Spezifikationen zu bewerten.
Sechs Experimente wurden entwickelt, um die Funktionalität der Gaia-X Credentials und des Datenaustausch-Workflows zu überprüfen: (1) Erlangung der Konformität für Teilnehmer-Credentials, (2) Erlangung der Konformität für Service-Offering-Credentials, (3) Erlangung der Konformität für Datenprodukt-Credentials, (4) Austausch von Fragebogendaten, (5) Austausch von Withings-Daten, und (6) Datenaustausch mit Verhandlung.

Das Scheitern der Experimente Nr. 2 und 3 deutet auf eine fehlende Implementierung und eine unvollständige Anpassung der implementierten Dienste an die definierten Spezifikationen hin.
Die Unfähigkeit, Experiment Nr. ~6 durchzuführen, zeigt Probleme mit fehlenden Implementierungen von Diensten, die für den Datenaustausch entscheidend sind.
Daher wird der Austausch medizinischer Studiendaten mit Gaia-X in der derzeitigen aktiven Entwicklungsphase als nicht durchführbar angesehen.

\vspace{5.0cm}

\begin{center}
\bfseries
Abstract
\normalfont
\end{center}

Data spaces are a data integration concept in which the data is not stored in central data silos, but the data is kept stored at the source.
Individual federations are set up to enable data sharing, where the data consumers are connected to data providers via interoperable standards.

Gaia-X aims to create data spaces in an \textit{interoperable}, \textit{transparent}, \textit{open}, and \textit{secure} environment that respects \textit{data rights} and \textit{EU regulations}.
Gaia-X Credentials are a central concept of the ecosystem.
They store descriptions of all participants and logical objects in a well-defined and verifiable fashion.
The EU trusts the project to decrease dependence on foreign tech giants and foster innovation in the digital economy.

The work investigates the feasibility of exchanging medical trial data in Gaia-X data spaces.
The thesis includes a literature review on data spaces and related topics.

The main method of evaluation is the development of a data exchange module for the existing platform for storing medical trial data---Carecentive---using Gaia-X components and specifications.
The exchange module acts as a proof of concept and is later used to conduct experiments to assess the working state of the Gaia-X services and specifications.
Six experiments were designed to verify the functionality of Gaia-X Credentials and the data exchange workflow: (1) obtaining compliance for Participant credentials, (2) obtaining compliance for Service Offering credentials, (3) obtaining compliance for Data Product credentials, (4) exchange of questionnaire data, (5) exchange of Withings data, and (6) data exchange with negotiation.

The failure of experiments no.~2 and 3 points to missing implementation and imperfect alignment of implemented services with defined specifications.
The inability to conduct experiment no.~6 demonstrates issues with missing implementations of services crucial for data exchange.
Therefore, the exchange of medical trial data using Gaia-X is deemed unfeasible in the current active development phase.


% Print a table fo contents
\tableofcontents
\cleardoublepage

% Nomenclature
% \addcontentsline{toc}{chapter}{Nomenclature}
%\null\newpage
%\printnomenclature[4em]
%\cleardoublepage

% start arabic page numbers
\pagenumbering{arabic}

% CONTENT
% Here we include the actual content of the thesis.
% Feel free to add/remove/rename the files and include statements according to your needs.
% Each file should start with a \chapter{Title} command
\include{contents/text}

% APPENDIX
% these files should appear in the appendix of the thesis
% you can again use as many or as few files and chapters as required for your thesis
\appendix
% \include{contents/app-01-example}   % Glossar (\chapter{Glossar}  TEXT)
% \include{contents/app-02-example}   %
% \include{contents/app-03-example}   %

% Don't change these includes
\include{includes/mad-lof} % List of Figures
\include{includes/mad-lot} % List of Tables
\include{includes/mad-lit} % Bibliography
\input{acronyms} % Acronym definition and list of acronyms

\end{document}
