\chapter{Methods}\label{ch:methods}

\begin{chapterabstract}
    The following chapter outlines the methodology used to assess the Gaia-X initiative, the reference framework implementation XFSC, and to answer the research questions stated in the introductory Chapter~\ref{ch:introduction}.
    The main method of work is implementation of a Gaia-X-compliant data exchange module, serving as a proof of concept.
    Based on this, the Gaia-X-related technical challenges are noted.
    Lastly, a series of experiments is presented as a way to ensure proper functionality of the XFSC services, followed with a section on its evaluation.
\end{chapterabstract}

\section{About Carecentive}\label{sec:about-carecentive}

Carecentive~\cite{carecentive} is a backend software framework used to support health studies and trials, which was developed at FAU MaD Lab.
The framework offers features that are typically required in medical trials, notably the following:
\begin{itemize}
    \item User management (registration, login, password management)
    \item Questionnaire storage \& submission
    \item Activity scheduling (e.g., periodic questionnaire submission)
    \item File upload (e.g., images of medical records)
    \item Wearable devices integration (Withings API, Fitbit API)
    \item User activity analytics
    \item Permission management (Roles such as patients, relatives, physicians, nurses, study managers)
\end{itemize}

\subsection{Tech Stack}\label{subsec:tech-stack}

From the developmental standpoint, the framework is divided into two packages, \textit{carecentive-framework\footnote{\url{https://github.com/carecentive/carecentive-framework}}} which mainly sets up the application, registers and exposes the API on the outside.
The API requests are then handed over to the main package \textit{carecentive-core\footnote{\url{https://github.com/carecentive/carecentive-core}}}, which handles user requests based on its application logic.

The Carecentive is written in \textit{JavaScript} and runs in the \textit{Node.js\footnote{\url{https://nodejs.org}}} engine.
It is built on top of the \textit{Express\footnote{\url{https://expressjs.com/}}} web framework, enabling the RESTful API interface.
It stores data in the relational \textit{MySQL\footnote{\url{https://www.mysql.com/}}} DBMS (Database Management System).
The database tables are versioned via database migrations enabled by the \textit{Knex.js\footnote{\url{https://knexjs.org/}}} library, and relational data is mapped onto objects (models) using the \textit{Objection.js\footnote{\url{https://vincit.github.io/objection.js/}}} library.

To simplify setting up the project on different machines and improve stability, the \textit{Docker\footnote{\url{https://www.docker.com/}}} platform was used.
This platform ensures the installation of all dependencies and runs the app in an isolated and reproducible environment, reducing the differences in various operating systems.

\subsection{Use Cases}\label{subsec:use-cases}

Currently, a smartphone app is used as a frontend for the Carecentive backend counterpart in a running study investigating the health of patients receiving care at the palliative care unit in the \textit{Uniklinikum Erlangen} (University Hospital Erlangen).
Thanks to the Carecentive framework's extensibility, other students' theses under the FAU MaD Lab also utilized it (e.g., Smartphone-based Urinalysis).

\section{Implementation}\label{sec:implementation}

In order to assess the functionality of the Gaia-X ecosystem, a partial goal of this thesis is to implement a Gaia-X-compliant module for the exchange of the data stored in the Carecentive app.
During the implementation of the exchange module, serving as a proof of concept, any technical or other issues with the Gaia-X specifications and the XFSC software are noted.
After the implementation is finished, a set of predefined data exchange scenarios is run to verify the correct functionality of data exchange module implementation and, mainly, whether the Gaia-X specifications and XFSC software are ready for real-world usage.
This is done within the specific use case of the palliative care trial.

The implementation task was to design and implement a workflow that would allow Carecentive users with the \texttt{admin} role to register the questionnaire and Withings\footnote{Withings is a French company developing smart wearable devices for measuring vital signs, activity, and sleep monitoring. Carecentive integrates the Withings API to download the data into its database.} data via Gaia-X-compliant processes to allow data consumption by users outside of Carecentive.
To be specific, the ``\texttt{/api/measurements}'' and ``\texttt{/api/questionnaires}'' API resources, which are protected by a temporary valid token issued upon user login, shall be able to accept a new type of token a non-Gaia-X user obtains upon successfully going through the Gaia-X contracting process.

To enable Carecentive users to share data in a Gaia-X-compliant manner, a collection of new resources was developed and exposed on the Carecentive API. Table~\ref{tab:endpoints} lists all provided endpoints with their description, authorization, and HTTP method; the ``\texttt{/api}'' prefix is common for all endpoints and is omitted for brevity.

\begin{longtable}{ |p{4cm}|p{2cm}|p{3cm}|p{6cm}| }
    \hhline{----}
    \textbf{Endpoint} & \textbf{HTTP Method} & \textbf{Authorization} & \textbf{Description}\\
    \hhline{----}
    \texttt{/admin/gaia-x/participants} & \textbf{\texttt{GET}} & Authenticated Admin & Fetches all registered Gaia-X Participants\\
    \hhline{----}
    \texttt{/admin/gaia-x/participants/:participantId} & \textbf{\texttt{GET}} & Authenticated Admin & Fetches a single Gaia-X Participant\\
    \hhline{----}
    \texttt{/admin/gaia-x/participants} & \textbf{\texttt{POST}} & Authenticated Admin & Registers a Gaia-X Participant and creates related credentials\\
    \hhline{----}
    \texttt{/gaia-x/data-products} & \textbf{\texttt{GET}} & Unrestricted & Fetches all registered data products\\
    \hhline{----}
    \texttt{/gaia-x/data-products/:dataProductId} & \textbf{\texttt{GET}} & Unrestricted & Fetches a single data product\\
    \hhline{----}
    \texttt{/gaia-x/data-products} & \textbf{\texttt{POST}} & Authenticated Admin & Registers a data product and creates related Gaia-X credentials\\
    \hhline{----}
    \texttt{/gaia-x/data-products/:dataProductId/contracts} & \textbf{\texttt{POST}} & Unrestricted & Creates a contract proposal for a data product\\
    \hhline{----}
    \texttt{/gaia-x/data-products/:dataProductId/contracts/:contractId} & \textbf{\texttt{PUT}} & Unrestricted & Publishes a signed contract for the given data product\\
    \hhline{----}
    \texttt{/gaia-x/data-products/:dataProductId/contracts/:contractId} & \textbf{\texttt{GET}} & Contract Signature & Fetches the signed contract from both parties\\
    \hhline{----}
    \texttt{/admin/gaia-x/data-product-contracts} & \textbf{\texttt{GET}} & Authenticated Admin & Fetches all data product contracts (proposals)\\
    \hhline{----}
    \texttt{/admin/gaia-x/data-product-contracts/:dataProductContractId} & \textbf{\texttt{GET}} & Authenticated Admin & Fetches a data product contract (proposal)\\
    \hhline{----}
    \texttt{/admin/gaia-x/data-product-contracts/:dataProductContractId} & \textbf{\texttt{PUT}} & Authenticated Admin & Signs a data product contract proposal\\
    \hhline{----}
    \texttt{/admin/gaia-x/data-product-contracts/:dataProductContractId} & \textbf{\texttt{DELETE}} & Authenticated Admin & Rejects a data product contract proposal\\
    \hhline{----}
    \texttt{/gaia-x/authentication} & \textbf{\texttt{POST}} & Unrestricted & Issues a temporary access token based on a valid contract\\
    \hhline{----}
    \caption{An overview of endpoints implemented into Carecentive along with the HTTP method used, authorization method and the description.}
    \label{tab:endpoints}
\end{longtable}

Now, let's look at the workflows of a Carecentive admin registering a participant and a data resource, contracting between the user and admin, and finally, the user's consumption of data workflow.

\subsection{Participant registration}\label{subsec:participant-registration}

Before a data product can be registered, the Carecentive admin needs to create a participant.
Registration is done by sending a \texttt{POST} request to the ``\texttt{/api/admin/gaia-x/participants}'' endpoint, specifying the details about the organization (\textit{VAT ID}, \textit{name} of the organization, \textit{country code} of the organization's legal address).
A certificate chain (must be of type EV SSL or eIDAS compliant) and a corresponding private key are required to sign the Gaia-X Credentials and create a DID.
Lastly, a so-called ``\texttt{participantSlug}'' is needed --- a unique alphanumerical string that serves mainly as a directory name for storing all participant-related files.

Figure~\ref{fig:participant_req} showcases a sample \texttt{POST} request to the ``\texttt{/api/admin/gaia-x/participants}'' endpoint.

\begin{figure}[h]
    \centering
    \begin{minted}[bgcolor=LightGray,breaklines,breakanywhere]{http}
POST /api/admin/gaia-x/participants HTTP/1.1
Content-Type: application/json

{
	"certificateChain": [
		"-----BEGIN CERTIFICATE-----\nMIIDXTCCAkWgAwIBAgIJAKH3BLaGxskwMA0GCSqGSIb3DQEBCwUAMEUxCzAJBgNV\n...\nJmCVzwB/MCuv3z4IV0IsbP5ZKh+6m0oI+RINdDl4Stxdj6LK8JspNeU0R6mBg1A=\n-----END CERTIFICATE-----",
		"-----BEGIN CERTIFICATE-----\nMIIDXTCCAkWgAwIBAgIJAMDOIDZf9l9yMA0GCSqGSIb3DQEBCwUAMEUxCzAJBgNV\n...\nv5pXVIlZm8DtRmDTUp8pNDnQ5UAZTuDPaWo3zbxZPOUmGJ1IlBOgaQvsyPCW5gU=\n-----END CERTIFICATE-----",
	],
	"privateKey": "-----BEGIN PRIVATE KEY-----\nMIIEvQIBADANBgkqhkiG9w0BAQEFAASCAmMwggJfAgEAAoGBAM6Q5O3vVfnxk6P7\n...\nYWw+HiFJh9XQpRUtv9PV8L8AqFFfMdsOpT6pgC+aA/WB\n-----END PRIVATE KEY-----",
	"vatId": "DE132507686",
	"organizationName": "Acme Corporation",
	"countryCode": "DE-BY",
	"participantSlug": "acme-corp"
}
    \end{minted}
    \caption{Example of a participant registration request.}\label{fig:participant_req}
\end{figure}

A successful request results in the creation of a database record.
The certificate chain is also stored and used to create a DID document.
The established DID document then enables the creation of the \textit{LegalRegistrationNumber}, \textit{TermsAndConditions}, and, finally, \textit{Participant} credentials.

\subsection{Data Product Registration}\label{subsec:data-product-registration}

Once a participant is created, it can be used for exposing data products under it.
This is achieved by sending a \texttt{POST} request to the ``\texttt{/api/gaia-x/data-products}'' endpoint, specifying the participant ID in Carecentive and the route to be shared (at the moment, this can be ``\texttt{/api/questionnaires}'' or ``\texttt{/api/measurements}'').
Following this, you must include the information used to create the Gaia-X credentials.
For the creation of credentials, the inclusion of \texttt{title}, \texttt{description}, \texttt{termsAndConditions}, \texttt{termsOfUsage}, \texttt{license}, and \texttt{policy} fields is mandatory.
The inclusion of \texttt{dataCreatedAt}, \texttt{dataExpiresAt}, and \texttt{dataLanguageCode} fields is optional.
Finally, the private key has to be included again since it is not stored on the server for security reasons.

Figure~\ref{fig:data_products_req} showcases a sample \texttt{POST} request to the ``\texttt{/api/gaia-x/data-products}'' endpoint.

\begin{figure}[h]
    \centering
    \begin{minted}[bgcolor=LightGray,breaklines,breakanywhere,breakanywhere]{http}
POST /api/gaia-x/data-products HTTP/1.1
Content-Type: application/json

{
	"participantId": 1,
	"title": "Palliative Care Questionnaires",
	"description": "sample description",
	"termsAndConditions": "sample terms and conditions",
	"termsOfUsage": "sample terms of usage",
	"license": "sample license",
	"policy": "default: allow",
	"route": "/api/questionnaires",
	"dataCreatedAt": "2024-07-01 12:00:00",
	"dataExpiresAt": "2025-01-01 12:00:00",
	"dataLanguageCode": "en",
	"privateKey": "-----BEGIN PRIVATE KEY-----\nMIIEvQIBADANBgkqhkiG9w0BAQEFAASCAmMwggJfAgEAAoGBAM6Q5O3vVfnxk6P7\n...\nYWw+HiFJh9XQpRUtv9PV8L8AqFFfMdsOpT6pgC+aA/WB\n-----END PRIVATE KEY-----"
}
    \end{minted}
    \caption{Example of a data product creation request.}\label{fig:data_products_req}
\end{figure}

This results in creating a database record, enabling anyone to browse the available data product.
Crucially, the following general Gaia-X credentials are created: \textit{ServiceOffering}, \textit{DataResource}, \texttt{SoftwareResource}, \texttt{InstantiatedVirtualResource}, \texttt{ServiceAccessPoint}, followed by \texttt{DataProductDescription}, \texttt{DatasetDescription}, and \texttt{DataUsage} credentials defined by the \textit{Data Exchange Services Specifications}~\cite{gaiax_data_exchange_document} document.

\subsection{Data Contracting}\label{subsec:data-contracting}

Contracting is initiated by the user sending a \texttt{POST} request to the ``\texttt{/api\-/gaia-x\-/data-products\-/:dataProductId\-/contracts}'' endpoint, which expects a JSON object with a single attribute \texttt{participantUrl}, which has to be a resolvable URL to the user's valid Participant Credential.
The backend fetches the Participant Credential, its issuer's DID document, and the DID's certificate.
Afterwards, the Participant Credential is validated and its signature is verified against the DID certificate.
Lastly, the server creates a contract proposal based on the selected data product and provided hashes of the provided participant, DID document, user certificate, and the contract proposal are stored to prevent being tampered with during the contracting phase.
The contract proposal record is then stored in the DB, and the proposal is returned to the user for signing.
With that, the contract proposal moves to the state ``\texttt{CONSUMER\_SIGNATURE\_PENDING}.''

Before the contracting can take place, the user first has to browse and select the desired data product.
The \texttt{GET} method on the ``\texttt{/api\-/gaia-x\-/data-products}'' endpoint serves this purpose.
Once the user has selected the desired data product, they need to take note of its ID in the form of a UUID\footnote{UUID is a Universally Unique Identifier designed to be unique across different systems. The UUIDv4, which we use, follows the form \texttt{xxxxxxxx-xxxx-Mxxx-Nxxx-xxxxxxxxxxxx}, where M determines the version (4) and the top 2 bits of N depict the variant used (10\textsubscript{2} in our case). The rest of the bits (122) are randomly generated.}, which is used to identify the chosen data product.

Now, the user has to sign the contract, by which they confirm agreement with the conditions of usage.
The signature has to be formatted as a JWS2020 (JSON Web Signature)\footnote{\url{https://www.w3.org/TR/vc-jws-2020/}}, which is stored in the form of three base64-encoded values separated by a dot --- \texttt{[header].[content].[signature]}.
This process follows the standard VC and Gaia-X signing scheme, where the header contains information about the used signing algorithm, and the VC content is canonized using the \texttt{URDNA2015\footnote{\url{https://www.w3.org/community/reports/credentials/CG-FINAL-rdf-dataset-canonicalization-20221009/}}} algorithm.
Then, a SHA256 hash is computed and signed with the keypair associated with the participant's DID.
Gaia-X then mandates the removal of the \texttt{[content]} part of the JWS (the hash value) since it's redundant as the signature is attached to the VC anyway.
The final format of the signature is the following: \texttt{[header]..[signature]}.

Once the contract proposal is signed by the user and attached under the ``\texttt{proof}'' attribute of the contract, they can upload it to the Carecentive API, where it will await signature from the admin.
This submission of the signed contract proposal is made by sending a \texttt{PUT} request to the endpoint ``\texttt{/api\-/gaia-x\-/data-products\-/:dataProductId\-/contracts\-/:dataProduct\-ContractId}'' with the signed contract attached as the request body.

The Carecentive backend re-checks the integrity of the Participant credential, DID document, user certificate, and contract contents.
After all checks are successfully passed, the contract signature is verified against the participant's certificate.
If the signature is verified successfully, the contract proposal moves to the state ``\texttt{PRODUCER\_SIGNATURE\_PENDING}.''

Now, the contract awaits signing by a Carecentive admin, who can do so via the PUT method on the ``\texttt{/api\-/admin\-/gaia-x\-/data-product-contracts\-/:dataProductContractId}'' endpoint.
This endpoint expects a JSON object with a single attribute, ``\texttt{privateKey},'' as the request body.
By providing a valid private key, the server co-signs the contract, and the state changes to ``\texttt{READY\_TO\_BE\_CLAIMED}.''

The data contract is now signed by both parties and is ready to be claimed by the user.
The user claims the contract by calling the ``\texttt{/api\-/gaia-x\-/data-products\-/:dataProductId\-/contracts\-/:contractId}'' endpoint using the \texttt{GET} method.
The user also has to provide his contract signature of the contract as a \textit{Bearer} token in the \textit{Authorization} header to prevent unauthorized access.

When the consumer claims the data contract, the contract moves to the state ``\texttt{FINALIZED},'' concluding the data exchange's contracting phase.

Before the data contract reaches the ``\texttt{FINALIZED}'' state, a Carecentive admin can also reject the contract proposal by calling the ``\texttt{/api\-/admin\-/gaia-x\-/data-product\--contracts\-/:dataProduct\-ContractId}'' endpoint using the \texttt{DELETE} method.
This action moves the contract to the ``\texttt{REJEC\-TED}'' state, the second terminal state, and an alternative way to conclude the data contracting phase.

\subsection{Data Consumption}\label{subsec:data-consumption}

Once the data consumer has a valid and signed data product contract, they can begin consuming the data.
To improve security and reduce the bandwidth of each request, the contract is not presented each time the user consumes data.
The process is divided into two steps:
\begin{enumerate}
    \item Access token issuance
    \item Data consumption
\end{enumerate}

The user first has to obtain a temporarily valid access token.
This is done by sending a \texttt{POST} request to the ``\texttt{/api\-/gaia-x\-/authentication}'' endpoint with the data contract attached as the request body.
The Carecentive server verifies the validity of the contract and issues an access token valid for 10 minutes provided the validation is successful.

The data consumer can then use the token to access the endpoint route specified in the data product (``\texttt{/api\-/questionnaires}'' or ``\texttt{/api\-/measurements}'').
To ensure successful authentication, the user must call the endpoint with the access token in the \texttt{Authorization} header as a \texttt{Bearer} token.

\subsection{Code Organization}\label{subsec:code-organization}

To ensure proper maintainability of the Carecentive project, the data exchange module follows the structure established by the existing project.
The convention includes a logical separation of code into specific types of files, namely:

\begin{itemize}
    \item Models --- Map database rows onto specific objects.
    \item Services --- Handle complex application logic.
    \item Routes --- Act as an entry-point for a request (they perform tasks usually handled by Controllers in an \textit{MVC} or \textit{MVP} architecture).
    \item Migrations --- Denote changes in the database structure to allow database versioning.
\end{itemize}

To improve code readability, the data exchange component introduces two new class types:
\begin{itemize}
    \item Helpers --- Handle a specific subset of the problem handled by one or more Services.
    \item Validators --- Enforce validation of the request body.
\end{itemize}

Let's go over the specific classes that handle the application logic --- \textit{Services} and \textit{Helpers}:
\begin{itemize}
    \item \texttt{ThirdPartyTokenService} --- Handles the issuance of access tokens to enable Carecentive access to users outside the platform.
    \item \texttt{ParticipantStorage} --- A Helper that handles the storage of Gaia-X Credentials, DID documents, and other related files in the specific Participant-allotted storage space.
    \item \texttt{DidService} --- Handles the creation and management of DID.
    It is supported by the \texttt{ParticipantStorage} helper.
    \item \texttt{GaiaXService} --- Handles the creation, signing, and management of Gaia-X Credentials.
    It uses the \texttt{DidService} to get a Participant's DID and the \texttt{ParticipantStorage} helper to handle credential storage.
    \item \texttt{ParticipantService} --- Handles the creation of new Participants.
    It uses the \texttt{DidService} for DID creation, the \texttt{GaiaXService} for the creation of Participant Gaia-X Credentials, and the \texttt{ParticipantStorage} for the cleanup of Participant data.
    \item \texttt{DataProductService} --- Handles the creation of new data products and associated Gaia-X Credentials.
    It uses the \texttt{GaiaXService} to issue Gaia-X Credentials and the \texttt{ParticipantStorage} to store related non-credential data.
    \item \texttt{DataProductContractService} --- Handles the creation and verification of data product contracts (proposals) and issuance of temporary access tokens.
    Additionally, it handles the verification of contracts and user-supplied artifacts (Participant Credential, DID document).
    It uses the \texttt{GaiaXService} for Credential-related operations, the \texttt{ThirdPartyTokenService} for issuing access tokens, and the \texttt{ParticipantStorage} helper for reading Participant's certificate for producer signature verification.
\end{itemize}

Jest\footnote{\url{https://jestjs.io/}}, a popular JavaScript testing framework, was used for automated testing to discover potential issues and ensure correct behavior when introducing new changes in the code.

\section{Experiments}\label{sec:experiments}

To judge the proper functioning of the Carecentive data-exchange module, services of the GXDCH, reference Gaia-X implementation, and Gaia-X Credentials themselves, a series of experiments were conducted:
\begin{enumerate}
    \item Obtaining Gaia-X compliance for Participant-related Credentials
    \item Obtaining Gaia-X compliance for ServiceOffering-related Credentials
    \item Obtaining Gaia-X compliance for DataProduct-related Credentials
    \item Exchange of questionnaire data
    \item Exchange of Withings data
    \item Data exchange with policy negotiation
\end{enumerate}

Below is detailed information about each of the experiments.

\subsection[Participant Compliance]{Obtaining Gaia-X compliance for Participant-related Credentials}\label{subsec:participant-compliance}

The test is performed by calling the \texttt{POST} method on the Carecentive's ``\texttt{/api/admin/gaia-x/participants}'' endpoint using the data in Figure~\ref{fig:test_case_1}:

\begin{figure}[h]
    \centering
    \begin{minted}[bgcolor=LightGray,breaklines,breakanywhere]{http}
POST /api/admin/gaia-x/participants HTTP/1.1
Content-Type: application/json

{
	"certificateChain": [
		"-----BEGIN CERTIFICATE-----\nMIIDXTCCAkWgAwIBAgIJAKH3BLaGxskwMA0GCSqGSIb3DQEBCwUAMEUxCzAJBgNV\n...\nJmCVzwB/MCuv3z4IV0IsbP5ZKh+6m0oI+RINdDl4Stxdj6LK8JspNeU0R6mBg1A=\n-----END CERTIFICATE-----",
		"-----BEGIN CERTIFICATE-----\nMIIDXTCCAkWgAwIBAgIJAMDOIDZf9l9yMA0GCSqGSIb3DQEBCwUAMEUxCzAJBgNV\n...\nv5pXVIlZm8DtRmDTUp8pNDnQ5UAZTuDPaWo3zbxZPOUmGJ1IlBOgaQvsyPCW5gU=\n-----END CERTIFICATE-----",
	],
	"privateKey": "-----BEGIN PRIVATE KEY-----\nMIIEvQIBADANBgkqhkiG9w0BAQEFAASCAmMwggJfAgEAAoGBAM6Q5O3vVfnxk6P7\n...\nYWw+HiFJh9XQpRUtv9PV8L8AqFFfMdsOpT6pgC+aA/WB\n-----END PRIVATE KEY-----",
	"vatId": "DE132507686",
	"organizationName": "Acme Corporation",
	"countryCode": "DE-BY",
	"participantSlug": "acme-corp"
}
    \end{minted}
    \caption{Request for the experiment test case no.~1.}\label{fig:test_case_1}
\end{figure}

The request is performed using the eIDAS-compliant qualified certificate for electronic signature and a valid FAU's VAT~ID.
The certificate chain and private key are redacted for privacy reasons.
The request leads to the creation of a DID document, \texttt{TermsAndConditions}, and \texttt{Participant} Gaia-X Credentials.
The \texttt{LegalRegistrationNumber} credential is obtained from the Gaia-X Notary by performing the request shown in Figure~\ref{fig:test_case_1_notary}.

\begin{figure}[h]
    \centering
    \begin{minted}[bgcolor=LightGray,breaklines]{http}
POST https://registrationnumber.notary.lab.gaia-x.eu/v1/registrationNumberVC HTTP/1.1
Content-Type: application/json

{
  "@context": [
    "https://registry.lab.gaia-x.eu/development/api/trusted-shape-registry/v1/shapes/jsonld/participant"
  ],
  "type": "gx:legalRegistrationNumber",
  "id": "https://gaia-x-dev.simerda.dev/gaia-x/acme-corp/legal-registration-number.json#cs",
  "gx:vatID": "DE132507686"
}
    \end{minted}
    \captionsetup{width=\textwidth}
    \caption{LegalRegistrationNumber request to the Gaia-X Notary service.}\label{fig:test_case_1_notary}
\end{figure}

A response with HTTP code 200 and a validated \texttt{LegalRegistrationNumber} credential signed by the Notary Trust Anchor in the response body is expected.

Next, a request to the Gaia-X Compliance service is made by performing the request shown in Figure~\ref{fig:test_case_1_compliance}, where the \texttt{LegalRegistrationNumber}, \texttt{TermsAndConditions}, and \texttt{Participant} placeholders are replaced with the actual respective credentials.

\begin{figure}[h]
    \centering
    \begin{minted}[bgcolor=LightGray,breaklines,escapeinside=||]{http}
POST https://compliance.lab.gaia-x.eu/v1/api/credential-offers?vcid=https://gaia-x-dev.simerda.dev/gaia-x/acme-corp/compliance.json HTTP/1.1
Content-Type: application/json

{
  "@context": "https://www.w3.org/2018/credentials/v1",
  "type": "VerifiablePresentation",
  "verifiableCredential": [
    {|LegalRegistrationNumber|},
    {|TermsAndConditions|},
    {|Participant|}
  ]
}
    \end{minted}
    \caption{Participant compliance request to the Gaia-X Compliance service}\label{fig:test_case_1_compliance}
\end{figure}

A response with HTTP code 201 and a compliance VC issued by the Compliance Trust Anchor in the response body is expected.

Upon successfully passing all steps, the Carecentive API responds with the HTTP code 201 and returns the created participant model.

\subsection[Service Offering Compliance]{Obtaining Gaia-X compliance for ServiceOffering-related Credentials}\label{subsec:service-offering-compliance}

Since this test relies on the participant record, the possibility of executing this experiment depends on successfully completing test no.~1.
The test is performed by calling the \texttt{POST} method on the Carecentive's ``\texttt{/api\-/gaia-x\-/data-products}'' endpoint using the data in Figure~\ref{fig:test_case_2}:

\begin{figure}[h]
    \centering
    \begin{minted}[bgcolor=LightGray,breaklines,breakanywhere]{http}
POST /api/gaia-x/data-products HTTP/1.1
Content-Type: application/json

{
    "participantId": 1,
	"title": "Questionnaires",
	"description": "sample questionnaire description",
	"termsAndConditions": "sample terms and conditions",
	"termsOfUsage": "sample terms of usage",
	"license": "sample license",
	"policy": "default: allow",
	"route": "/api/questionnaires",
	"dataCreatedAt": "2024-07-01 12:00:00",
	"dataExpiresAt": "2025-01-01 12:00:00",
	"dataLanguageCode": "en",
	"privateKey": "-----BEGIN PRIVATE KEY-----\nMIIEvQIBADANBgkqhkiG9w0BAQEFAASCAmMwggJfAgEAAoGBAM6Q5O3vVfnxk6P7\n...\nYWw+HiFJh9XQpRUtv9PV8L8AqFFfMdsOpT6pgC+aA/WB\n-----END PRIVATE KEY-----"
}
    \end{minted}
    \caption{Request for the experiment test case no.~2.}\label{fig:test_case_2}
\end{figure}

The request is performed using a private key corresponding to the certificate used for participant with ID 1, created in the previous test.
The private key is redacted for privacy reasons.
The request leads to the creation of \texttt{ServiceOffering}, \texttt{DataResource}, \texttt{SoftwareResource}, \texttt{InstantiatedVirtualResource}, \texttt{ServiceAccessPoint}, \texttt{DataProductDescription}, \texttt{DatasetDescription}, and \texttt{DataUsage} Gaia-X Credentials and the \texttt{license}, \texttt{terms-and-conditions} and \texttt{terms-of-usage} files.

Upon successful execution, the Carecentive API shall respond with the HTTP code 201 and return the created data product model with links to all created Gaia-X Credentials.

Next, a request to the Gaia-X Compliance service is made by performing the request similar to the one carried out in test no.~1, shown in Figure~\ref{fig:test_case_1_compliance}.
Contrary to the previous test, the \texttt{ServiceOffering}, \texttt{DataResource}, \texttt{SoftwareResource}, \texttt{InstantiatedVirtualResource}, and \texttt{ServiceAccessPoint} Gaia-X Credentials are packed into the VP body.

A response with HTTP code 201 and a compliance VC issued by the Compliance Trust Anchor in the response body is expected.

\subsection[Data Product Compliance]{Obtaining Gaia-X compliance for DataProduct-related Credentials}\label{subsec:data-product-compliance}

Since this test relies on the participant record, the possibility of executing this experiment depends on successfully completing test no.~1.
The test is performed by calling the \texttt{POST} method on the Carecentive's ``\texttt{/api\-/gaia-x\-/data-products}'' endpoint using the data in Figure~\ref{fig:test_case_3}:

\begin{figure}[h]
    \centering
    \begin{minted}[bgcolor=LightGray,breaklines,breakanywhere]{http}
POST /api/gaia-x/data-products HTTP/1.1
Content-Type: application/json

{
    "participantId": 1,
	"title": "Withings",
	"description": "sample withings description",
	"termsAndConditions": "sample terms and conditions",
	"termsOfUsage": "sample terms of usage",
	"license": "sample license",
	"policy": "default: allow",
	"route": "/api/measurements",
	"dataCreatedAt": "2024-08-01 14:32:51",
	"dataExpiresAt": "2026-02-13 20:15:41",
	"dataLanguageCode": "en",
	"privateKey": "-----BEGIN PRIVATE KEY-----\nMIIEvQIBADANBgkqhkiG9w0BAQEFAASCAmMwggJfAgEAAoGBAM6Q5O3vVfnxk6P7\n...\nYWw+HiFJh9XQpRUtv9PV8L8AqFFfMdsOpT6pgC+aA/WB\n-----END PRIVATE KEY-----"
}
    \end{minted}
    \caption{Request for the experiment test case no.~3.}\label{fig:test_case_3}
\end{figure}

The request is performed using a private key corresponding to the certificate used for the participant with ID 1, created in test no.~1.
The private key is redacted for privacy reasons.
The request leads to the creation of \texttt{ServiceOffering}, \texttt{DataResource}, \texttt{SoftwareResource}, \texttt{InstantiatedVirtualResource}, \texttt{ServiceAccessPoint}, \texttt{DataProductDescription}, \texttt{DatasetDescription}, and \texttt{DataUsage} Gaia-X Credentials and the \texttt{license}, \texttt{terms-and-conditions} and \texttt{terms-of-usage} files.

Upon successful execution, the Carecentive API shall respond with the HTTP code 201 and return the created data product model with links to all created Gaia-X Credentials.

Next, a request to the Gaia-X Compliance service is made by performing the request similar to the one performed in test no.~1, shown in Figure~\ref{fig:test_case_1_compliance}.
Contrary to the first test, the \texttt{DataProductDescription}, \texttt{DatasetDescription}, and \texttt{DataUsage} Gaia-X Credentials are packed into the VP body.

A response with HTTP code 201 and a compliance VC issued by the Compliance Trust Anchor in the response body is expected.

\subsection{Exchange of questionnaire data}\label{subsec:exchange-of-questionnaire-data}

Since this test relies on the \textit{questionnaire} data product record, the possibility of executing this experiment depends on the successful completion of at least the first phase of test no.~2 (call to the Carecentive API).

The experiment is executed as outlined in the Sections~\ref{subsec:data-contracting} and~\ref{subsec:data-consumption}.

The experiment is considered successful upon passing the following steps:
\begin{enumerate}
    \item Successful contracting resulting in a contract signed by the producer and consumer parties.
    \item Successful issuance of an access token upon presenting the contract.
    \item Successful fetch of the questionnaire data.
\end{enumerate}

\subsection{Exchange of Withings data}\label{subsec:exchange-of-withings-data}

Since this test relies on the \textit{Withings} data product record, the possibility of executing this experiment depends on the successful completion of at least the first phase of test no.~3 (call to the Carecentive API).

The experiment is executed as outlined in the Sections~\ref{subsec:data-contracting} and~\ref{subsec:data-consumption}.

The experiment is considered successful upon passing the following steps:
\begin{enumerate}
    \item Successful contracting resulting in a contract signed by the producer and consumer parties.
    \item Successful issuance of an access token upon presenting the contract.
    \item Successful fetch of the Withings data.
\end{enumerate}

\subsection{Data exchange with policy negotiation}\label{subsec:data-exchange-with-policy-negotiation}

Since this test relies on the participant record, the possibility of executing this experiment depends on successfully completing test no.~1.
The test is performed by calling the \texttt{POST} method on the Carecentive's ``\texttt{/api/gaia-x/data-products}'' endpoint using the data in Figure~\ref{fig:test_case_6}, replacing the \texttt{\{policy\}} placeholder with policy defined in Figure~\ref{fig:test_case_6_policy}:

\begin{figure}[h]
    \centering
    \begin{minted}[bgcolor=LightGray,breaklines,breakanywhere]{http}
POST /api/gaia-x/data-products HTTP/1.1
Content-Type: application/json

{
    "participantId": 1,
	"title": "Questionnaires",
	"description": "sample questionnaire description",
	"termsAndConditions": "sample terms and conditions",
	"termsOfUsage": "sample terms of usage",
	"license": "sample license",
	"policy": "{policy}",
	"route": "/api/questionnaires",
	"dataCreatedAt": "2024-07-01 12:00:00",
	"dataExpiresAt": "2025-01-01 12:00:00",
	"dataLanguageCode": "en",
	"privateKey": "-----BEGIN PRIVATE KEY-----\nMIIEvQIBADANBgkqhkiG9w0BAQEFAASCAmMwggJfAgEAAoGBAM6Q5O3vVfnxk6P7\n...\nYWw+HiFJh9XQpRUtv9PV8L8AqFFfMdsOpT6pgC+aA/WB\n-----END PRIVATE KEY-----"
}
    \end{minted}
    \caption{Request for the experiment test case no.~6.}\label{fig:test_case_6}
\end{figure}

The request is performed using a private key corresponding to the certificate used for participant with ID 1, created in the first test.
The private key is redacted for privacy reasons.
The request creates \textit{ServiceOffering}, \textit{DataResource}, \textit{SoftwareResource}, \textit{InstantiatedVirtualResource}, \textit{ServiceAccessPoint}, \textit{DataProductDescription}, \textit{DatasetDescription}, and \textit{DataUsage} Gaia-X Credentials and the \textit{license}, \textit{terms-and-conditions} and \textit{terms-of-usage} files.

\begin{figure}[h]
    \centering
    \begin{minted}[bgcolor=LightGray,breaklines]{json}
{
 "@context": "https://www.w3.org/ns/odrl.jsonld",
 "@type": "Set",
 "uid": "https://gaia-x-dev.simerda.dev/gaia-x/acme-corp/241de422-e7af-4ed4-8e30-2f6f07726b19/policy.json",
 "permission": [
   {
     "target": "https://gaia-x-dev.simerda.dev/api/questionnaires",
     "assignee": "Data Consumer",
     "action": "use",
     "duty": [
       {
         "assigner": "Data Provider",
         "assignee": "Data Consumer",
         "action": [
           {
             "value": "compensate",
             "refinement": [
               {
                 "leftOperand": "payAmount",
                 "operator": "eq",
                 "rightOperand": { "@value": "100.00", "@type": "xsd:decimal" },
                 "unit": "https://dbpedia.org/resource/Euro"
               }
             ]
           }
         ]
       }
     ]
   }
 ],
 "obligation": [
   {
     "target": "https://gaia-x-dev.simerda.dev/api/questionnaires",
     "assigner": "Data Consumer",
     "assignee": "Data Provider",
     "action": "distribute"
   }
 ]
}
    \end{minted}
    \caption{The policy used for the experiment test case no.~6, obliging the data provider to allow access to data upon the payment of 100~\char"20AC~.}\label{fig:test_case_6_policy}
\end{figure}

Upon successful execution, the Carecentive API shall respond with the HTTP code 201 and return the created data product model with links to all created Gaia-X Credentials.

The experiment is considered successful upon passing the following steps:
\begin{enumerate}
    \item Rejection of the policy by consumer
    \item Counter-proposal of a policy requiring the payment of 50~\char"20AC~ for access to the data
    \item Acceptance of the counter-proposal by the data provider
    \item Successful contract signing by the producer and consumer
    \item Successful issuance of an access token upon presenting the contract
    \item Successful fetch of the questionnaire data
\end{enumerate}

\section{Evaluation}\label{sec:evaluation}

The conducted experiments are evaluated into three predefined categories, which represent the outcome of each experiment:

Pass --- The experiment was successful, passing all the test steps and resulting in the expected outcome as defined in Section~\ref{sec:experiments}.
Fail --- The experiment failed due to one or more steps failing or the outcome differing from the expected outcome as defined in Section~\ref{sec:experiments}.
Unable to Conduct --- The experiment couldn't be conducted due to unsatisfied preconditions.

If the experiment results in the \textit{Fail} outcome, a comment on the reasons for failure is provided.
If the experiment results in the \textit{Unable to Conduct} outcome, the preconditions that weren't satisfied and led to the inability to conduct the experiment are provided.
They can be one of the following:
\begin{itemize}
    \item Failure of a prerequisite test --- The test depends on another test, which failed.
    \item Missing XFSC component --- The test uses a component from the XFSC reference, which is not available.
    \item GXDCH services unreachable --- The test uses a service that is part of the GXDCH, and none of the instances were running when the experiment was conducted.
\end{itemize}

The results of these tests are presented in the following chapter.
